---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/philmikejones/rakeR.svg?branch=master)](https://travis-ci.org/philmikejones/rakeR)
[![codecov](https://codecov.io/gh/philmikejones/rakeR/branch/master/graph/badge.svg)](https://codecov.io/gh/philmikejones/rakeR)


rakeR
=====

Create a spatial microsimulated data set in R using iterative proportional 
fitting ('raking').


Install
-------

Install the development version with `devtools`:

```{r install-rakeR, eval=FALSE, include=TRUE}
# Obtain devtools if you don't already have it installed
# install.packages("devtools")

# Install rakeR development version from GitHub
devtools::install_github("philmikejones/rakeR")
```

Load the package with:

```{r attach-rakeR}
library("rakeR")
```


Raking
------

To perform the raking you should supply two data frames, one with the constraint
information with counts per category for each zone (e.g. census counts) and one
with individual--level data (i.e. one row per individual).
In addition supply a character vector with constraint variable names.

```{r data}
cons <- data.frame(

  "zone"   = letters[1:3],
  "a0_49"  = c(8, 2, 7),
  "a_gt50" = c(4, 8, 4),
  "f"      = c(6, 6, 8),
  "m"      = c(6, 4, 3)

)

inds <- data.frame(

  "id"     = LETTERS[1:5],
  "age"    = c("a_gt50", "a_gt50", "a0_49", "a_gt50", "a0_49"),
  "sex"    = c("m", "m", "m", "f", "f"),
  "income" = c(2868, 2474, 2231, 3152, 2473),
  stringsAsFactors = FALSE

)

vars <- c("age", "sex")
```

* (Re-)weighting is done with `weight()` which returns a data frame of 
fractional weights.
* Integerisation is performed with `integerise()` which returns a data frame of
integerised weights.
* `simulate()` takes care of creating the final microsimulated data and returns
a data frame of simulated cases in zones.

These functions can be combined with pipes:

```{r chain-example}
# obtain magrittr if not already installed
# install.packages("magrittr")
library("magrittr")

sim_df <- weight(cons, inds, vars) %>% integerise() %>% simulate(inds = inds)
head(sim_df)
```

Alternatively use the `rake()` function, which is a wrapper for
`weight() %>% integerise() %>% simulate()`:

```{r rake-exaple}
sim_df <- rake(cons, inds, vars)
head(sim_df)
```


Contributions
-------------

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

Feedback on the API, bug reports/issues, and pull requests are very welcome.

If making a pull request **do not** push to `origin/master`!
`origin/master` is a protected branch and expects CI tests to have been
successfully completed before it will merge code.

Develop in a new branch, check your changes with `devtools::check()`, and 
submit a pull request which will be checked by Travis CI.


Acknowledgements
----------------

Many of the functions in this package are based on code written by 
[Robin Lovelace](https://github.com/Robinlovelace) and 
[Morgane Dumont](https://github.com/modumont) for their book 
[*Spatial Microsimulation with R* (2016), Chapman and Hall/CRC Press](https://www.crcpress.com/Spatial-Microsimulation-with-R/Lovelace-Dumont/p/book/9781498711548). Their book is also an excellent resource for learning 
about spatial microsimulation and understanding what's going on under the hood
of this package.

The rewighting (ipfp) algorithm itself is 
[written by Andrew Blocker](https://github.com/awblocker/ipfp) 
and is written in `C` for maximum speed and efficiency.

Thanks to [Tom Broomhead](http://mhs.group.shef.ac.uk/members/tom-broomhead/)
for feedback on the function naming.


Contact
-------

philmikejones at gmail dot com


Licenses
--------

Copyright 2016 Phil Mike Jones.

rakeR is free software: you can redistribute it and/or modify
it under the terms of the GNU General Public License as published by
the Free Software Foundation, either version 3 of the License, or
(at your option) any later version.

rakeR is distributed in the hope that it will be useful,
but WITHOUT ANY WARRANTY; without even the implied warranty of
MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
GNU General Public License for more details.

You should have received a copy of the GNU General Public License
along with rakeR. If not, see <http://www.gnu.org/licenses/>.

See [LICENSE](https://github.com/philmikejones/rakeR/blob/master/LICENSE)
